---
output:
  pdf_document: default
  html_document: default
---


<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. 
-->



```{r setup_ch500, include=FALSE}

# packages

```


# Clean data principles {#clean-data}

placeholder for chapter 5


* The principles of data quality

* Understanding what makes data “dirty”, and why context matters

* Understanding tidy data principles

* Serving the research question



[@De_Jonge_van_der_Loo_2013]

[@Hubbard_data_cleaning_2017]


## Why clean data matters

* accuracy

* ability to link to other data sources—consistency matters



## Data Quality

[@Ligon_1996]

[@Ligon_2015], [@Ligon_2007]

### Data quality literature

There is quite a lot of literature on how to define data quality; the concepts in the hierarchical dimensions described by Wang, Reddy and Kon ^[Richard Y. Wang, M.P. Reddy and Henry B. Kon, "Toward quality data: An attribute-based approach", _Decision Support Systems_, 13 (1995) 349-372] is as follows:

* Accessible

  - Available
  
* Interpretable

  - Syntax
  
  - Semantics
  
* Useful

  - Relevant
  
  - Timely
  
    - Current
    
    - Non-volatile
    
* Believable

  - Complete
  
  - Consistent
  
  - From a credible source
  
  - Accurate
  

"Dirty data" is data that falls short on the _believable_ dimension, in particular evaluating whether the data are complete, consistent, and accurate. 

**Complete**

For our purposes, "complete" means whether any values in each record are missing (internally complete).

Is it possible to determine whether the records are an accurate representation of the whole. With a well-designed sample, it is possible that a sub-set of the population can provide an accurate set of measures about the population. 

**Consistent**

We will consider a measure to be consistent if the same value is reported in the same way.

Some examples:

* units are consistent; temperature is consistently reporting in degrees Celsius, not mixing Farenheit and Kelvin, or income is in a single currency.

* spelling is consistent. "British Columbia", "B.C." and "BC". Or the 57 different ways to spell "Philadelphia". ^[In this tweet, you can see 57 different ways of spelling "Philadelphia" in the data collected in a US Government loan form (https://twitter.com/dataeditor/status/1280278987797942272?s=20)]

**Accurate**

(How to evaluate?)




If the data fails to meet our standards or quality, we need to _clean the data_. Which doesn't sound like a lot of fun. Didn't we want to be data scientists or business intelligence experts or academic scientists uncovering the insights hiding in the data? 

> The act of cleaning data imposes values/judgments/interpretations upon data intended to allow downstream analysis algorithms to function and give results. That’s exactly the same as doing data analysis. In fact, “cleaning” is just a spectrum of reusable data transformations on the path towards doing a full data analysis. —Randy Au, 2020-09-15 ^[["Data Cleaning IS Analysis, Not Grunt Work"](https://counting.substack.com/p/data-cleaning-is-analysis-not-grunt)]


So what are we doing when we say we are "cleaning the data"? And how can we confirm that it is "cleaned" in the way that we have defined?


**The first challenge**: how do we find these things?

**The second challenge**: what can and should we do about them?


See [@Greve_Clean_data]


## Structure

The first place to start with cleaning your data is the structure.


### Tidy data

Tidy data\index{tidy data}

[@tidydata]



### Row counts (records)

Did your import step give you as many rows as you anticipated?


### Columns (variables)

Did your import step capture all of the variables?

Did the variables import as the correct classes?

Things to look for:

* did variables that should be numbers load as numeric types, or as character?

* if working with labelled variables, did they load as factor type?



We've read in our raw data. The next step is to make sure that the data is what we expect, and what we need for our analysis. That is, what is the data's _quality_? 

A different way this is sometimes approached is whether or not the data is _dirty_ and needs _cleaning_.

What are common dirty data problems?

* Column headers are values instead of variable names, or there are duplicate variable names.

* Multiple data types in a column. For example, some records are numbers and some are character strings—and this could be entirely correct, but if some number entry includes commas as thousands separators, it might be stored as a character.

* Multiple variables are stored in one column (examples include first name & last name, city & province or state).

* Misspellings (typing errors) 

* Value in wrong field (e.g. country entered into `city` field).

* Irregularities in unit of measurement (in an international survey or form, `salary` completed by respondents in the values of their local currency).

* Contradiction (for example, two databases with the same person but the date of birth differs, perhaps due to non-ISO8601 entry: 07-08-79 and 08-07-79 both have the same digits but one could be mm-dd-yy and the other dd-mm-yy...we just don't know which is the correct one. Or is one a typo?)

* Default values in the place of missing values.

> Dr Davis Lawrence, director of safety-literature database the SafeLit Foundation...tells me that 'in most US states the quality of police crash reports is at best poor for use as a research tool. ... Data-quality checks were rare and when quality was evaluated it was found wanting. For example, in Louisiana for most crashes in the 1980s most of the occupants were males who were born on January 1st, 1950. Almost all of the vehicles involved in crashes were the 1960 model year.' Except they weren't. These were just the default settings. [@Criado_Perez_2019, p.190]

* Values are transformed automagically by software. Microsoft's Excel spreadsheet program is the most famous of these, converting many non-date values into date format, and assigning new values to store the data. For example, entering the character string "MAR1" gets converted to March 1st of the current year. 

* Dates may be stored as a numeric value (or "serial number") representing the number of days elapsed from a fixed starting point—a starting point that varies by operating system. For example, Excel for Windows uses January 1, 1900 as the first day, while earlier versions Macintosh computers, and by extension Excel for Macintosh, used January 1, 1904 as the start date. Thus the same date would be stored as different values. ^["Differences between the 1900 and the 1904 date system in Excel"](https://docs.microsoft.com/en-us/office/troubleshoot/excel/1900-and-1904-date-system)

```{r}
readxl::read_excel("date.xlsx", col_types = "text")

lubridate::today() - 44571
```


* Other (I'm not sure how to characterize this—wikipedia uses the term "cultural ignorance" which might fit)

> "Prawo Jazdy" was a supposed Polish national who was listed by the Garda Síochána in a police criminal database as having committed more than 50 traffic violations in Ireland. A 2007 memorandum stated that an investigation revealed prawo jazdy [ˈpra.vɔ ˈjaz.dɨ] to be Polish for 'driving licence', with the error arising due to officers mistaking the phrase, printed on Polish driving licences, to be a personal name while issuing traffic tickets. [@wiki:prawo_jazdy]


See [@Zumel_Mount_2019, chapters 3 and 4]






## Exploratory data analysis

One way to identify dirty data is through the process of _exploratory data analysis_ (EDA). This consists of two groups of techniques:

* summary statistics

* data visualization

### Summary statistics

(need to revise data set)

```{r}

# read in the data
customer_data <- readRDS("data_temp/custdata.RDS")

```

We can use R's `summary()` function to quickly get a glimpse of the data.

The things to look for:

* missing values

* invalid values and outliers

* data ranges that are too wide or too narrow

* the units of the data

```{r}
summary(customer_data)
```

### Data visualization

A plot can show us a wealth of information that cannot be expressed in the summary statistics. Here's the `age` variable from the customer data:

```{r}

summary(customer_data$age)

```


And here is the plot:

```{r}

ggplot(customer_data, aes(x = age)) +
  geom_density(bw = .75)

```






## Validation {#validation-intro}



## Feature engineering

Somehow and somewhere, the word "feature" has come to replace "variable" in the lexicon of the data scientist.

"Feature engineering" is a jazzed-up term meaning "calculating a new variable from others in the data"...for example, the unemployment rate in any given month and location is calculated by dividing the number of unemployed people by the total number of people in the labour force (two "features" in our data).

Similarly, "dimensionality reduction" means "selecting the variables to include in your analysis (or model)".

[@R-bcEpiRate]

